OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-06-08 main.tex
Included Marc's comments

\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$,
François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb
1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
Cuartero\,$^{\text{\sfb 1,2}}$, Thomas Graf\,$^{\text{\sfb 1,2}}$, Marc A.
Mart\'i Renom\,$^{\text{\sfb 1,2}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$
and Guillaume J. Filion\,$^{\text{\sfb 1,2}}$}

\address{$^{\text{\sf 1}}$Gene Regulation, Stem Cells and Cancer Program,
Centre for Genomic Regulation (CRG), The Barcelona Institute of Science
and Technology (BIST), Barcelona, Spain \\
$^{\text{\sf 2}}$Universitat Pompeu Fabra (UPF), Barcelona, Spain}

@@

One of the most significant breakthroughs towards this insight was the
development of the high throughput chromosomal conformation capture
technology (Hi-C), assaying chromosomal contacts at a genome-wide scale
\citep{lieberman2009comprehensive}. Nowadays, exploring the spatial
organization of chromatin has become a priority in many fields and Hi-C
has become part of the standard molecular biology toolbox
\citep{denker2016second}.

Contrary to the precursor technologies 3C, 4C and 5C \citep{de2012decade},
Hi-C interrogates all possible pairwise interactions between restriction
fragments. However, this does not guarantee that the method has no bias.
On the contrary, local genome features such as the G+C content, the
availability of restriction enzyme sites and the mappability of the
sequencing reads have been shown to impact the results
\citep{yaffe2011probabilistic}. In addition, general experimental biases
such as batch effects exist. It is thus important to normalize Hi-C data in
order to remove biases and artifacts, so that they are not confused with
biological signal.

@@

Once the parameters of the model are estimated, we rescale the estimated
means $\{\lambda_i\}$ to obtain a correction vector $\{\lambda_i'\}$ that
can be used to compute the corrected counts ($\hat{x}_{ij}$).

@@

The model is fitted with the Baum-Welch algorithm \citep{baum1966} until
convergence, following the detail given by \cite{filion2010systematic}.
The Viterbi path is then computed and corresponds to the inferred copy
number of each bin ($c_i$).

@@

To test the correction of biases, we gathered a set of
published \citep{ledily2014distinct, encode2012integrated, rao20143d,
stadhouders2017transcription, lin2012global, dixon2012topological} and
unpublished Hi-C data of different cell types and organisms. The details
can be found in Table \ref{tab:samples}.

@@

We also used array-based copy-number segmentation of the two cell
lines obtained from COSMIC database \citep{forbes2010cosmic}.

@@

We developed the routines contained in the \texttt{dryhic} R package to
efficiently create sparse representations of contact matrices and further
apply \textit{vanilla}, \textit{ICE} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. All the results are based on a
resolution of 100 kbp, but we found no major differences for different
values (not shown).

@@

Experimental biases affect the total amount of contacts in a continuous
but not necessarily linear way. Figure~\ref{fig:totals}A shows the
relationship between the amount of contacts and the number of restriction
enzyme sites in T47D, a breast cancer cell line with aberrant and unstable
karyotype.  Four clouds are visible. Each corresponds to sequences present
in one to four copies. In all of them, the relationship flattens as the
number of restriction sites increases. In order to capture this
relationship, \textit{OneD} fits a non-linear model between the total
amount of contacts and the known sources of bias (by default the G+C
content, the number of restriction sites and the mappability of the
reads).

@@

We tested the validity of this approach against the Catalogue Of Somatic
Mutations In Cancer \citep[COSMIC,][]{forbes2010cosmic}. In what follows,
we benchmarked \textit{OneD}, against \textit{vanilla}, \textit{ICE}
\citep{imakaev2012iterative}, \textit{caICB} \citep{wu2016computational}
and the Local Genomic Features method
\citep[\textit{LGF},][]{hu2012hicnorm, servant2012hitc}. The first three
methods correct biases implicitly, whereas the fourth method does it
explicitly. Figure \ref{fig:copy_number} shows the Pearson correlation
between the corrected number of contacts and the copy number estimation
for both T47D and K562 cell lines. Similar results were obtained using
Spearman correlation (Supplementary Figure 1). All the methods except
\textit{OneD} decrease the agreement between the signal and the copy
number because they partially correct the discrepancy. In contrast,
\textit{OneD} enhances the conformity of the signal with the copy number.
Not correcting for variable copy number at that stage may seem
counter-intuitive, but the tests below will show this leads to better
performance.

\begin{figure}[b]
	\centerline{\includegraphics[width=.45\textwidth]
{img/copy_number_figure2.pdf}}
\caption{Pearson correlation between the corrected number of contacts per bin
and an independent estimation of the copy number (COSMIC). Left panel
T47D breast cancer cell line, right panel K562 leukemia cell line.}
\label{fig:copy_number}
\end{figure}

@@

We first benchmarked the performance of our approach on biological
samples with an aberrant karyotype. A good normalization method should
increase the similarity between biological replicates by reducing
irrelevant experimental variation, such as batch effects, laboratory of
origin and protocol variations. Similarly, a good normalization should
decrease the similarity between samples that received a different
treatment in order to enhance the biological differences.

We assembled two Hi-C data sets obtained from T47D and K562, two cancer
cell lines with aberrant karyotypes. In each set we spiked two samples
from the other cell line (see Table~\ref{tab:samples}) in order to
introduce biological variability. We compared matrices before and after
normalization by different methods using the Pearson correlation of
observed over expected counts (see \ref{sec:comp}). This gave an
indication of the impact of a given normalization method. The results are
summarized in Figure~\ref{fig:aberrant}.

@@

An important application of normalization methods in experimental setups
is to identify outliers. We thus investigated the capacity of the
different methods to help identify the samples from the other cell type
spiked in the data set. We interpreted the pairwise comparison scores as
classifier scores and summarized the results with a ROC curve
(Figures~\ref{fig:aberrant}C and D).

All the methods, including the absence of normalization, succeeded in
identifying the T47D outliers among the K562 samples, but recognizing the
K562 outliers among the T47D samples proved more challenging.
\textit{OneD} increased the discrimination power compared to the raw
matrices, but all the other methods decreased it. Actually, they performed
little better than random on this task.

Using the other metrics described in \ref{sec:comp} yielded similar
results (Supplementary Figures 5, 6 and 7). Note that Spearman correlation
of contacts presented the worst performance for all scenarios, and it thus
seems to be a poor choice as a metric to compare Hi-C matrices.

@@

When interpreting the similarity scores as classification scores, we
observed that all the methods could identify approximately 50\% of the
biological pairs, after which their performance diverged
(Figure~\ref{fig:normal}B). \textit{OneD} achieved the highest area under
the curve on this problem, but with a small margin over the other methods
except \textit{vanilla}.

Using other metrics to compare matrices gave similar results
(Supplementary Figures 7 and 8). \textit{OneD} was always among the top
scoring methods. In these conditions, Spearman correlation of contacts
again appeared as the worst comparison metric because it showed a lower
performance for all the normalization methods.

@@

Removing the explicit experimental biases enhances the importance of the
copy number in the signal (Figure~\ref{fig:copy_number}). The copy number
is not an experimental bias, but it may be considered an additional source
of bias to be remove. For this reason, \textit{OneD} also allows the user
to correct the copy number and output a euploid-equivalent matrix.

Here, \textit{OneD} segments the linear signal of the total amount of
contacts into piecewise homoegenous regions. This is carried out by a
hidden Markov model whereby the averages of the states are constrained to
be within a discrete factor of each other. This allows the model to detect
regions with a number of copies equal to 1, 2, 3 \textit{etc}. With the
inferred number of copies at hand, \textit{OneD} then normalizes each cell
of the matrix with equation (\ref{eq:cnvnorm}).

Figure~\ref{fig:cnv_correction} shows the effect of this process on the
final Hi-C matrix. Panel \ref{fig:cnv_correction}A indicates that
different normalization methods can either enhance or diminish the signal
in regions with higher copy number. In this example, \textit{ICE}
overcompensates the original bias at the center of the picture and fades
the signal almost entirely. Concomitantly, the signal at the bottom left
of the matrix is enhanced and shows a structure that is not visible in the
raw data. On the contrary, \textit{LGF} strengthen the central region and
the diagonal. \textit{OneD} reduces the level of the central portion by a
factor 2 approximately, but does not otherwise distort the main features
of the region. This example shows that the copy number does not have a
simple and predictable effect on the final matrix. Not taking it into
account may open the door to some normalization artifacts. The result
provided by \textit{OneD} is not necessarily the right one (see
Discussion), but at least it does not correct copy number variations
as a side effect of some other criteria.

@@

In summary, \textit{OneD} is the only method that can be used to obtain a
euploid-equivalent normalized matrix in cases that the effect of the copy
number must be removed from the signal.

@@

Finally, we compared the computational speed of the different
normalization methods. One of the main reasons for the broad acceptance of
\textit{vanilla} and \textit{ICE} as standards is the speed of the
available implementation \citep{imakaev2012iterative}. This is even more
significant as current explicit methods \citep{servant2012hitc} are much
slower in comparison.

@@

The originality of \textit{OneD} lies in that it projects all the biases
onto a single variable: the total amount of contacts per bin. This allows
greater running speed, while preserving a good performance on samples with
karyotypic aberrations. One of the reasons that \textit{OneD} is better
able to highlight the biological distinctions between samples is that it
only corrects the copy number if specifically requested by the user. The
impact of copy number variations on normalization is rather opaque,
especially if they are treated as implicit biases
(Figure~\ref{fig:cnv_correction}). Treating them as explicit biases with
optional removal seems to be an overall safer strategy.

This raises the question whether variations of the copy number constitute
a biological signal or an artifact. If the biological sample contains
karytoypic aberrations, then its genome is grossly different from the reference
so there can be no correct way to normalize the Hi-C signal. The proper
approach would be to use the actual genome of the biological sample as a
reference to construct the contact map. However, this strategy is
presently unfeasible because assembling mammalian genomes is still a hard
problem.

>>>>

\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$,
François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb
1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
Cuartero\,$^{\text{\sfb 1,2}}$, Thomas Graf\,$^{\text{\sfb 1,2}}$, Marc A.
Mart\'i Renom\,$^{\text{\sfb 1,2,3,4}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$
and Guillaume J. Filion\,$^{\text{\sfb 1,2}}$}

\address{$^{\text{\sf 1}}$Gene Regulation, Stem Cells and Cancer Program,
Centre for Genomic Regulation (CRG), The Barcelona Institute of Science
and Technology (BIST), Dr. Aiguader 88, 08003, Barcelona, Spain \\
$^{\text{\sf 2}}$Universitat Pompeu Fabra (UPF), Barcelona, Spain \\
$^{\text{\sf 3}}$CNAG-CRG, Centre for Genomic Regulation (CRG), Barcelona
Institute of Science and Technology (BIST), Baldiri i Reixac 4, 08028 Barcelona,
Spain \\
$^{\text{\sf 4}}$ICREA, Pg. Lluís Companys 23, 08010 Barcelona, Spain}

@@

One of the most significant breakthroughs towards this insight was the
development of the high throughput chromosomal conformation capture
technology (Hi-C), assaying chromosomal contacts at a genome-wide scale
\citep{lieberman2009comprehensive}. Nowadays, exploring the spatial
organization of chromatin has become a priority in many fields and Hi-C
has become part of the standard molecular biology toolbox
\citep{dekker2013exploring}.

Contrary to the precursor technologies 3C, 4C and 5C \citep{dekker2002capturing,simonis2006nuclear,dostie2006chromosome,de2012decade},
Hi-C interrogates all possible pairwise interactions between restriction
fragments. However, this does not guarantee that the method has no bias.
On the contrary, local genome features such as the G+C content, the
availability of restriction enzyme sites and the mappability of the
sequencing reads have been shown to impact the results
\citep{yaffe2011probabilistic}. In addition, general experimental biases
such as batch effects exist. It is thus important to normalize Hi-C data in
order to remove biases and artifacts, so that they are not confused with
biological signal.

@@

Once the parameters of the model are estimated, estimated
means $\{\lambda_i\}$ are rescaled to obtain a correction vector $\{\lambda_i'\}$ that
can be used to compute the corrected counts ($\hat{x}_{ij}$).

@@

The model is fitted with the Baum-Welch algorithm \citep{baum1966} until
convergence, following a previously described protocol \citep{filion2010systematic}.
The Viterbi path is then computed and corresponds to the inferred copy
number of each bin ($c_i$).

@@

To test the correction of biases, we gathered a set of
published \citep{ledily2014distinct, encode2012integrated, rao20143d,
stadhouders2017transcription, lin2012global, dixon2012topological} and
unpublished Hi-C data of different cell types and organisms
(Table \ref{tab:samples}).

@@

We also used array-based copy-number segmentation of the two cell
lines obtained from the COSMIC database \citep{forbes2010cosmic}.

@@

We developed the routines contained in the \texttt{dryhic} R package to
efficiently create sparse representations of contact matrices and further
apply \textit{vanilla}, \textit{ICE} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. All the results are based on a
resolution of 100 kbp, but we found no major differences at different
resolutions (not shown).

@@

Experimental biases affect the total amount of contacts in a continuous
but not necessarily linear way. Figure~\ref{fig:totals}A shows the
relationship between the amount of contacts and the number of restriction
enzyme sites in T47D, a breast cancer cell line with aberrant and unstable
karyotype.  Four clouds are visible. Each corresponds to sequences present
in one to four copies. In all of them, the relationship flattens as the
number of restriction sites increases. To capture this
relationship, \textit{OneD} fits a non-linear model between the total
amount of contacts and the known sources of bias (by default the G+C
content, the number of restriction sites and the mappability of the
reads).

@@

% Marc's comment here was regarding lack of motivation and explanation for the
% copy number comparison. I've rescued some text from an older version. We
% should also comment here about CN being a bias or providing biological
% information, so to correct or not to correct for CN

In what follows, we benchmarked \textit{OneD},
against \textit{vanilla}, \textit{ICE} \citep{imakaev2012iterative}, \textit{caICB} \citep{wu2016computational}
and the Local Genomic Features method \citep[\textit{LGF},][]{hu2012hicnorm,
servant2012hitc}. The first three methods correct biases implicitly, whereas the
fourth method does it explicitly.

Given that our model is based on total number of contacts, we reasoned that a preliminary test
would be to check if the corrected number of contacts per bin reflects the copy
number (as measured by an independent technique such as array-based copy-number
segmentation). We tested the validity of this approach against the Catalogue Of Somatic
Mutations In Cancer \citep[COSMIC,][]{forbes2010cosmic}. Figure \ref{fig:copy_number} shows the Pearson correlation
between the corrected number of contacts and the copy number estimation
for both T47D and K562 cell lines. Similar results were obtained using
Spearman correlation (Supplementary Figure 1). All the methods except
\textit{OneD} decrease the agreement between the signal and the copy
number because they partially correct the discrepancy. In contrast,
\textit{OneD} enhances the conformity of the signal with the copy number.
Not correcting for variable copy number at that stage may seem
counter-intuitive, but the tests below will show this leads to better
performance.

\begin{figure}[b]
	\centerline{\includegraphics[width=.45\textwidth]
{img/copy_number_figure2.pdf}}
\caption{Pearson correlation between the total number of contacts per bin
and an independent estimation of the copy number (COSMIC). Left panel
T47D breast cancer cell line, right panel K562 leukemia cell line.}
\label{fig:copy_number}
\end{figure}

@@

We first benchmarked the performance of our approach on biological
samples with an aberrant karyotype. A good normalization method should
increase the similarity between biological replicates by reducing
irrelevant experimental variation, such as batch effects, laboratory of
origin and protocol variations. Similarly, a good normalization should
decrease the similarity between samples that received a different
treatment to enhance the biological differences.

We assembled two Hi-C data sets obtained from T47D and K562, two cancer
cell lines with aberrant karyotypes. In each set we spiked two samples
from the other cell line (Table~\ref{tab:samples}) to
introduce biological variability. We compared matrices before and after
normalization by different methods using the Pearson correlation of
observed over expected counts (see \ref{sec:comp}). This gave an
indication of the impact of a given normalization method. The results are
summarized in Figure~\ref{fig:aberrant}.

@@

An important application of normalization methods in experimental setups
is to identify outliers. We thus investigated the capacity of the
different methods to help identify the samples from the other cell type
spiked in the data set. We interpreted the pairwise comparison scores as
classifier scores and summarized the results with a ROC curve
(Figures~\ref{fig:aberrant}C and D). All the methods, including the absence of
normalization, succeeded in identifying the T47D outliers among the K562
samples, but recognizing the K562 outliers among the T47D samples proved more
challenging. \textit{OneD} increased the discrimination power compared to the
raw matrices, but all the other methods decreased it. Actually, they performed
little better than random on this task. Using the other metrics described
in \ref{sec:comp} yielded similar results (Supplementary Figures 5, 6 and
7). Note that Spearman correlation of contacts presented the worst performance
for all scenarios, and it thus seems to be a poor choice as a metric to compare
Hi-C matrices.

@@

When interpreting the similarity scores as classification scores, we
observed that all the methods could identify approximately 50\% of the
biological pairs, after which their performance diverged
(Figure~\ref{fig:normal}B). \textit{OneD} achieved the highest area under
the curve on this problem, but with a small margin over the other methods
except \textit{vanilla}. Using other metrics to compare matrices gave similar
results (Supplementary Figures 7 and 8). \textit{OneD} was always among the top
scoring methods. In these conditions, Spearman correlation of contacts again
appeared as the worst comparison metric because it showed a lower performance
for all the normalization methods.

@@

Removing the explicit experimental biases enhances the importance of the
copy number in the signal (Figure~\ref{fig:copy_number}). The copy number
is not an experimental bias, but it may be considered an additional source
of bias to be remove. For this reason, \textit{OneD} also allows the user
to correct the copy number and output a euploid-equivalent matrix. To do
so, \textit{OneD} segments the linear signal of the total amount of contacts
into piecewise homoegenous regions. This is carried out by a hidden Markov model
whereby the averages of the states are constrained to be within a discrete
factor of each other. This allows the model to detect regions with a number of
copies equal to 1, 2, 3 \textit{etc}. With the inferred number of copies at
hand, \textit{OneD} then normalizes each cell of the matrix with equation
(\ref{eq:cnvnorm}).

Different normalization methods can either enhance or diminish the signal
in regions with higher copy number (Figure~\ref{fig:cnv_correction}). In this
example, \textit{ICE} overcompensates the original bias at the center of the
picture and fades the signal almost entirely. Concomitantly, the signal at the
bottom left of the matrix is enhanced and shows a structure that is not visible
in the raw data. On the contrary, \textit{LGF} strengthen the central region and
the diagonal. \textit{OneD} reduces the level of the central portion by a factor
2 approximately, but does not otherwise distort the main features of the
region. This example shows that the copy number does not have a simple and
predictable effect on the final matrix. Not taking it into account may open the
door to some normalization artifacts. The result provided by \textit{OneD} is
not necessarily the right one (see Discussion), but at least it does not correct
copy number variations as a side effect of some other criteria.

@@

In summary, \textit{OneD} can be used to obtain a euploid-equivalent normalized
matrix in cases that the effect of the copy number must be removed from the signal.

@@

Finally, we compared the computational speed of the different
normalization methods. \textit{vanilla} and \textit{ICE} have broad acceptance
for their conceptual simplicity, easy use and
speed \citep{imakaev2012iterative}. This is even more significant as current
explicit methods \citep{servant2012hitc} are much slower in comparison.

@@

The originality of \textit{OneD} lies in that it projects all the biases
onto a single variable: the total amount of contacts per bin. This allows
greater running speed, while preserving a good performance on samples with
karyotypic aberrations. One of the reasons why \textit{OneD} is able to better
highlight the biological distinctions between samples is that it only corrects
the copy number if specifically requested by the user. The impact of copy number
variations on normalization is rather opaque, especially if they are treated as
implicit biases (Figure~\ref{fig:cnv_correction}). Treating them as explicit
biases with optional removal seems to be an overall safer strategy.

This raises the question whether variations of the copy number constitute
a biological signal or an artifact. If the biological sample contains
karytoypic aberrations, then its genome is grossly different then the reference
genome, which makes signal correction very challenging. The proper approach
would be to use the actual genome of the biological sample as a reference to
construct the contact map. However, this strategy is presently unfeasible
because assembling mammalian genomes is still a hard problem.
